\textls[-0]{\thetitle}

### 2023-03-05 canonicitymodel.tex
Working on appendices.

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''.\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
We can even argue that, if this theorem is proven in a computable meta-logic\footnote{QIIT does have computational content~\cite{altkap2016}}, 
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.\YZ{Can we say something about strong/weak normalization?}\EDJreply{
  No. \\
  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
  4. Of course we can conjecture we have that normalization function \\
  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
  but we don't have termination (a runnable interpreter) for open terms.  \\ 
  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
}

@@

The proof can be found in appendix. The intuition of the model is similar to that of the consistency model---use the canonicity model of sigma type (from the earlier work) for linkage type and using Agda's inductive facility to justify our W-type.

>>>>

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''.\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
We can even argue that, if this theorem is proven in a computable meta-logic, 
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.\YZ{Can we say something about strong/weak normalization?}\EDJreply{
  No. \\
  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
  4. Of course we can conjecture we have that normalization function \\
  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
  but we don't have termination (a runnable interpreter) for open terms.  \\ 
  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
}

@@

% The proof can be found in appendix. 
The intuition of the model is similar to that of the consistency model---use the canonicity model of sigma type (from the earlier work) for linkage type and using Agda's inductive facility to justify our W-type.

### 2023-03-05 metatheory2.tex
Adding Info on appendices.
Change makefile
Working on appendices.

\subsection{Warm Up: Review Declarative MLTT}
We start with introduction to MLTT as \TT can be roughly considered as MLTT with linkages.

\input{syntax}

% Copy and Paste Intuition of MLTT from other places to here
\subsection{\TT : MLTT with Linkages}

>>>>

\subsection{Warm Up: Review Declarative MLTT}
We begin with a review of MLTT as \TT can be roughly considered as MLTT with linkages.
We work in an intrinsic-typed setting, so there is no un-typed preterms.

\input{all-mltt-typing}

% Copy and Paste Intuition of MLTT from other places to here
\subsection{\TT : an MLTT variant with Linkages}
Compared to MLTT, \TT has 3 more judgements about signatures and linkage transformers.

\input{all-fmltt-typing}

### 2023-03-05 standardmodel.tex
Working on appendices.

\newcommand{\goodCtxS}[2]{{ {#1} \ \models_S }}
\newcommand{\goodTypeS}[3]{{ {#1} \models_S {#2} }}
\newcommand{\goodTermS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSubS}[3]{{ {#1} \models_S {#2} : {#3} }}
\newcommand{\goodSigS}[3]{{ {#1} \models_S {#2} \ \  \texttt{LSig}^{#3} }}
\newcommand{\goodWSigS}[3]{{ {#1} \models_S {#2} \ \ \texttt{WSig}^{#3} }}
\newcommand{\goodSealS}[4]{{ {#1} \models_S {#2} : {#3} \  |\  {#4} }}
\newcommand{\goodInhS}[4]{{ {#1} \models_S {#2} : {#3} \twoheadrightarrow {#4}}}

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of QIIT. However, in this section, for readability, we only inductively interpret each syntax piece into naive set theory that is also respecting judgemental equalities. Please refer to our pseduo-Agda formulation for a more formal and complete version.

@@

The main intuition of the consistency model is that: (1) since we have used sigma type to translate the linkage, using the model for sigma type for linkage will work; (2) directly using the inductive facility of (pseudo)-Agda (or naive set theory) to justify the intro and elim rule of our W-type.

Note that, since every piece is intrinsic well-typed, $\denotesS{t}$ means the same as $\denotesS{\goodTerm{\Gm}{t}{T}}$ (i.e. every term is tracking its context and type).\footnote{And our (pseudo) Agda formulation is well-typed as well. Latex-formulation should be well-typed but we discarded all the type information in the paper.}

Here we will only show some interpretation and also omit the universe level to sketch out the idea, please refer to appendix for the complete version.

@@

  \denotesS{\goodCtx{\Gm, A}{i}} &= \{(\Gm, a) : \Gm \in \denotesS{\goodCtx{\Gm}{}}, a \in \denotesS{\goodType{\Gm}{A}{}}(\Gm)\}

@@

  \denotesS{\goodTerm{\Gm}{\_}{T}} & = \prod_{\Gm \in \denotesS{\goodCtx{\Gm}{i}}}\denotesS{T}(\Gm) \\

@@

  \denotesS{\goodType{\Gm}{\Sigma A B}{}}(\Gm) &:= \sum_{a \in \denotesS{A}\Gm} \denotesS{B}(\Gm, a) \\

@@

  \denotesS{\goodType{\Gm}{\cL \sigma}{i}} (\Gm)&= \{(m, t) :  m \in \denotesS{\cL \tau}(\Gm) \land t \in \prod_{m' \in \denotesS{A}(\Gm)}\denotesS{T}(\Gm, m')  \}  \\

@@

  \denotesS{\goodType{\Gm}{\cL \sigma}{i}} (\Gm)&= \{\star\} \text{ when } \denotesS{\sigma} \in \denotesS{\goodSig{\Gm}{\_}{0}} \\

@@

  \denotesS{\goodTerm{\Gm}{\LkgEmp}{\cL \LSigEmp} } (\Gm) &= \star \\
  \denotesS{\goodTerm{\Gm}{\LkgAdd{o}{t}}{\cL(\LSigAdd {\sigma}{s}  {T})}} \ (\Gm)&= (o(\Gm), (\lambda \ m' . t'(\Gm, m'))) \\
  \denotesS{\goodTerm{\Gm}{\lkgproj{1}{t}}{\cL(\lsigproj{1}{\sigma})}}(\Gm) &=  "fst"~(t\Gm) \\ 

@@

  \denotesS{\goodType{\Gm}{\bot}{i}} \ (\Gm) &= \emptyset \\
  \denotesS{\goodType{\Gm}{\cB}{i}} \ (\Gm) &= \{0, 1\} \quad \text{a specific two element set} \\

@@

We encourage the reader to read the (pseudo-)Agda-style proof with more type annotation so it is clearer about the mathematical object doing the interpretation. Here, we need to clarify some ambiguity in our latex formulation. For example, when we wrote ``$\denotesS{\goodType{\Gm}{\TyTkg {\sigma}}{}}$ defined inductively on ${\sigma}$'', the best way to understand it is to consider we are defining a function $\denotesS{\goodType{\Gm}{\TyTkg {\_}}{}}$ as the interpretation for $\goodType{\Gm}{\TyTkg {\_}}{}$ that mapping $\denotesS{\sigma}$ to $\denotesS{\goodType{\Gm}{\TyTkg {\sigma}}{}}$, inductively on the length of the signature. The case is similar for $\denotesS{\cL \_}$, $\denotesS{\Tkg {\_}}$ and  $\denotesS{\Sigma \ \_ \ \_}$. \textbf{This is actually the formulation in Agda}---QIIT syntax consider $\TyTkg{} : "Sig"~\Gm~n \to "Ty"~\Gm$ as a constructor and our model needs to interpret this constructor to a function $\TyTkg{}^S : "Sig"^S \Gm^S~n \to "Ty"^S~\Gm^S$ with each judgement also interpreted.

Basically in Agda, we are simply constructing a function that mapping our QIIT data-type to some other data---it maps $"Con"$ to $"Con"^S$, "Ty : Con → Set" to "Ty$^S$ : Con$^S$ → Set" and etc. What's more, since judgemental equality in our meta-theory is represented by equality in QIIT, and the well-definedness of this mapping-out function implies that ``judgementally equal syntax is interpreted by the same semantic in the model''.

@@

We sketch the model for the inductive type and inheritance judgement here, and both of them require inductive facility in meta-logic. In set theory, that would be using \textit{Rule sets}~\cite{timany2017consistency,aczel1998relating}, but we are using Agda's Inductive facility instead. Both of them are somewhat ``interpreted by themselves'' in Agda, because their syntactic definition is pretty ``well-founded''. For inductive type, for each grounded signature, we have an inductive type using introduction rule of W-type as constructor, shown in \cref{fig:model-ind-type}. This "MWsup" when applying to $n$ will use the $n$-th constructor specify in the signature. This grounded-ness is required to verify the substitution laws. For inheritance judgement, we simply construct a term model of inheritance judgement using vanilla inductive type (instead of QIIT). Then we use meta-level substitution to model the explicit substitution, and we can define $\denotesS{"inh"~\_ ~\_}$ by induction on this term model.

@@

The verification of validity of this model is largely omitted here and we refer the reader to the appendix. To verify our model preserve the judgemental equality, usually unpacking the definition is enough---for example, to verify $\beta$-rule for linkage projection, we only need to see that $\denotesS{\goodTerm{\Gm}{\lkgproj{1}{(\LkgAdd{o}{t})}}{\cL\sigma}} = \denotesS{\goodTerm{\Gm}{o}{\cL\sigma}}$ by function extensionality. One exception is for $\TyTkg{}$ and $\goodSig{\_}{\_}{n}$, the substitution law around these two are also proved together by induction because they were mutual-recursively defined together. The compatibility lemmas can be verified directly---for example $\frac{\goodSigS{\Gm}{\sigma}{n+1}\quad\goodTermS{\Gm}{o}{\cL\sigma}}{\goodTermS{\Gm}{(\lkgproj{1}{o})}{\cL(\lsigproj{1}{\sigma})}}$ is proved by simply unfolding the definition.

Since all the compatibility lemmas hold (by simply changing $\vdash$ into $\models$ in the syntactic rules then we would have all the compatibility lemmas\footnote{Apparently, in our (pseudo) Agda formulation, we don't need this extra step.}), we have constructed a model.
Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.\YZ{Explicitly state the theorem.}

The full model formulated in (pseudo) Agda syntax can be found in the appendix.

>>>>

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from~\cite{altkap2016,kaposi2017type, kaposi2019gluing} and their formalization, where they formally resort to the concept of \textit{algebra/model} of Quotient Inductive Inductive Type (QIIT). However, in our formulation, we only inductively interpret each syntax piece into naive set theory that is also respecting judgemental equalities.

@@

The main intuition of the consistency model is that: (1) since we have used sigma type to translate the linkage, using the model for sigma type for linkage will work; (2) directly using the inductive facility of naive set theory to justify the intro and elim rule of our W-type.

Note that, since every piece is intrinsic well-typed, $\denotesS{t}$ means the same as $\denotesS{\goodTerm{\Gm}{t}{T}}$ (i.e. every term is tracking its context and type).

Here we will only show the interpretation for  omit the universe level to sketch out the idea. 
% only show some interpretation and also

@@

  \denotesS{\goodCtx{\Gm, A}{i}} &= \{(\Gm, a) : \gm \in \denotesS{\goodCtx{\Gm}{}}, a \in \denotesS{\goodType{\Gm}{A}{}}(\gm)\}

@@

  \denotesS{\goodTerm{\Gm}{\_}{T}} & = \prod_{\gm \in \denotesS{\goodCtx{\Gm}{i}}}\denotesS{T}(\gm) \\

@@

  \denotesS{\goodType{\Gm}{\Sigma A B}{}}(\gm) &:= \sum_{a \in \denotesS{A}\Gm} \denotesS{B}(\Gm, a) \\

@@

  \denotesS{\goodType{\Gm}{\cL \sigma}{i}} (\gm)&= \{(m, t) :  m \in \denotesS{\cL \tau}(\gm) \land t \in \prod_{m' \in \denotesS{A}(\gm)}\denotesS{T}(\Gm, m')  \}  \\

@@

  \denotesS{\goodType{\Gm}{\cL \sigma}{i}} (\gm)&= \{\star\} \text{ when } \denotesS{\sigma} \in \denotesS{\goodSig{\Gm}{\_}{0}} \\

@@

  \denotesS{\goodTerm{\Gm}{\LkgEmp}{\cL \LSigEmp} } (\gm) &= \star \\
  \denotesS{\goodTerm{\Gm}{\LkgAdd{o}{t}}{\cL(\LSigAdd {\sigma}{s}  {T})}} \ (\gm)&= (o(\gm), (\lambda \ m' . t'(\Gm, m'))) \\
  \denotesS{\goodTerm{\Gm}{\lkgproj{1}{t}}{\cL(\lsigproj{1}{\sigma})}}(\gm) &=  "fst"~(t\Gm) \\ 

@@

  \denotesS{\goodType{\Gm}{\bot}{i}} \ (\gm) &= \emptyset \\
  \denotesS{\goodType{\Gm}{\cB}{i}} \ (\gm) &= \{0, 1\} \quad \text{a specific two element set} \\

@@

% We encourage the reader to read the (pseudo-)Agda-style proof with more type annotation so it is clearer about the mathematical object doing the interpretation. 
Here, we need to clarify some ambiguity in our latex formulation. For example, when we wrote ``$\denotesS{\goodType{\Gm}{\TyTkg {\sigma}}{}}$ defined inductively on ${\sigma}$'', the best way to understand it is to consider we are defining a function $\denotesS{\goodType{\Gm}{\TyTkg {\_}}{}}$ as the interpretation for $\goodType{\Gm}{\TyTkg {\_}}{}$ that mapping $\denotesS{\sigma}$ to $\denotesS{\goodType{\Gm}{\TyTkg {\sigma}}{}}$, inductively on the length of the signature. The case is similar for $\denotesS{\cL \_}$, $\denotesS{\Tkg {\_}}$ and  $\denotesS{\Sigma \ \_ \ \_}$. 
% \textbf{This is actually the formulation in Agda}---QIIT syntax consider $\TyTkg{} : "Sig"~\Gm~n \to "Ty"~\Gm$ as a constructor and our model needs to interpret this constructor to a function $\TyTkg{}^S : "Sig"^S \Gm^S~n \to "Ty"^S~\Gm^S$ with each judgement also interpreted.

% Basically in Agda, we are simply constructing a function that mapping our QIIT data-type to some other data---it maps constructor $"Con"$ to $"Con"^S$, and constructor "Ty : Con → Set" to "Ty$^S$ : Con$^S$ → Set" and etc. What's more, since judgemental equality in our meta-theory is represented by equality in QIIT, and the well-definedness of this mapping-out function implies that ``judgementally equal syntax is interpreted by the same semantic in the model''.

@@

We sketch the model for the inductive type and inheritance judgement here, and both of them require inductive facility in meta-logic. In set theory, that would be using \textit{Rule sets}~\cite{timany2017consistency,aczel1998relating}, but here we informally use Agda's Inductive facility instead. Both of them are somewhat ``interpreted by themselves'' in Agda, because their syntactic definition is pretty ``well-founded''. For inductive type, for each grounded signature, we have an inductive type using introduction rule of W-type as constructor, shown in \cref{fig:model-ind-type}. This "MWsup" when applying to $n$ will use the $n$-th constructor specify in the signature. This grounded-ness is required to verify the substitution laws. For inheritance judgement, we simply construct a term model of inheritance judgement using vanilla inductive type (instead of QIIT). Then we use meta-level substitution to model the explicit substitution, and we can define $\denotesS{"inh"~\_ ~\_}$ by induction on this term model.

@@

The verification of validity of this model is largely omitted here. To verify our model preserve the judgemental equality, usually unpacking the definition is enough---for example, to verify $\beta$-rule for linkage projection, we only need to see that $\denotesS{\goodTerm{\Gm}{\lkgproj{1}{(\LkgAdd{o}{t})}}{\cL\sigma}} = \denotesS{\goodTerm{\Gm}{o}{\cL\sigma}}$ by function extensionality. One exception is for $\TyTkg{}$ and $\goodSig{\_}{\_}{n}$, the substitution law around these two are also proved together by induction because they were mutual-recursively defined together. The compatibility lemmas can be verified directly---for example $\frac{\goodSigS{\Gm}{\sigma}{n+1}\quad\goodTermS{\Gm}{o}{\cL\sigma}}{\goodTermS{\Gm}{(\lkgproj{1}{o})}{\cL(\lsigproj{1}{\sigma})}}$ is proved by simply unfolding the definition.

Since all the compatibility lemmas hold (by simply changing $\vdash$ into $\models$ in the syntactic rules then we would have all the compatibility lemmas), we have constructed a model.
Finally, notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.\YZ{Explicitly state the theorem.}

### 2023-03-05 syn-translate.tex
Working on appendices.

We use the local morphism (the same one for encoding the Sealing
judgements) to translate inheritance judgement. This is mainly due to
the existence of \ruleref{Inh} rule. Here we only typeset the overriding
and inheritance rule, but all of the translation for inh judgment are
induced by the $\beta$-rule of \ruleref{Inh}.

We omit all the equational rules ($\beta,\eta$ and substitution) here.
However, for coherence, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cC}$ above, we actually have to prove the substitution laws
$(\cC\sigma)[\tau]\equiv \cC(\sigma[\tau])$ and $\Sigr{\Delta}{n}[\tau]
\equiv \Sigr{\Gamma}{n}$ together. The details (of all the equational
rules) can be found in the appendices written in (pseudo-)Agda form. We
encourage the reader to read that (pseudo-)Agda-style proof because it
is more friendly to type theorists, easy to do type checking and
validate the proof and it is clearer on the mathematical object each
translation uses.

Ultimately, we have constructed a model for our QIIT-syntax using only a
``subpart'' of our QIIT-syntax -- without these linkages. It is also
possible to compile our $\goodWSig{\Gamma}{w}{n}$ and W-type into a more
conventional formulation, but the recursor will always require a sigma
type as handler and thus the elimination rule for our W-type is still
non-conventional.

@@

% the link for syntactic translation
% https://drive.google.com/file/d/1pMqn8DS4T4jiCubzk3HgMANjrfKpcGlI/view?usp=sharing

>>>>

We use the local morphism\cite{abbott2003category} (the same one for encoding the Sealing
judgements) to translate inheritance judgement. This is mainly due to
the existence of \ruleref{Inh} rule. 
% Here we only typeset the overriding
% and inheritance rule, but all of the translation for inh judgment are
% induced by the $\beta$-rule of \ruleref{Inh}.

We omit all the equational rules ($\beta,\eta$ and substitution) here.
However, for coherence, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cC}$ above, we actually have to prove the substitution laws
$(\cC\sigma)[\tau]\equiv \cC(\sigma[\tau])$ and $\Sigr{\Delta}{n}[\tau]
\equiv \Sigr{\Gamma}{n}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.

Ultimately, we have constructed a model for our syntax using only a
``subpart'' of our syntax -- without these linkages. It is also
possible to compile our $\goodWSig{\Gamma}{w}{n}$ and W-type into a more
conventional formulation, but the recursor will always require a sigma
type as handler and thus the elimination rule for our W-type is still
non-conventional.

@@
